Analyzing changes in Canadian Grocery Prices

### 2024-11-14 paper/paper.qmd
Update paper

---
title: "Analyzing changes in Canadian grocery prices"
# subtitle: "My subtitle if needed"
author: 
  - Abdullah Motasim
  - Elizabeth Luong
thanks: "Code and data are available at: [https://github.com/RohanAlexander/starter_folder](https://github.com/RohanAlexander/starter_folder)."
date: today
date-format: long
abstract: "This paper investigates grocery pricing trends across Canada using SQL analysis on the 'Project Hammer' dataset, which contains detailed vendor-specific price data. Focusing on statistical relationships and potential pricing biases, we explore factors driving price variations across different regions and vendors. A Bayesian regression model is applied to understand correlations, accounting for limitations and highlighting areas where causation cannot be assumed. Findings reveal competitive dynamics within Canadian grocery markets, implications for consumers, and recommendations for future research into market and pricing behavior across vendors."
format: pdf
number-sections: true
bibliography: references.bib
---

@@

library(tidyverse)
library(palmerpenguins)
library(ggplot2)
library(dplyr)
library(knitr)
library(rstanarm)
library(arrow)
library(data.table)
```

@@

The rising cost of groceries in Canada has drawn significant public attention, impacting consumer budgets and prompting regulatory interest. While inflation and supply disruptions are often cited as contributors, the specifics of how major Canadian grocery vendors set and adjust prices remain underexplored. This study addresses this gap by analyzing data from Project Hammer, a data set that tracks grocery prices across vendors including Voila, T&T, Loblaws, No Frills, Metro, Galleria, Walmart, and Save-On-Foods. Through examining these data, we aim to uncover factors driving grocery price dynamics and assess the degree to which vendor-specific pricing may reflect competitive or non-competitive practices.

@@

The paper is structured as follows: @sec-data discusses the data types included in the raw data, the cleaning process for the data, and the reason for selecting the data set we did. @sec-model discusses model specification and justification for utilizing a Bayesian linear regression model. @sec-results presents the trends and correlations between different variables utilizing tabular and graphical means. @sec-discussion discusses the results of @sec-results going into detail on what the simulation results can tell us about grocery prices in Canada as well as discussing missing data and sources of bias.

@@

For this study, we utilize data from Project Hammer, a Canadian initiative designed to monitor grocery prices across major retailers in Canada. The Project Hammer dataset was collected from eight prominent Canadian grocery vendors—Voila, T&T, Loblaws, No Frills, Metro, Galleria, Walmart, and Save-On-Foods—between February 28, 2024, and the latest available data load in the database [ citeProjectHammerData]. This dataset enables us to investigate price fluctuations, identify pricing trends, and explore potential competitive or collusive patterns within the Canadian grocery sector.
Project Hammer aims to support regulatory efforts by providing data on price transparency, making it suitable for academic research and potential legal analysis. Following data processing and cleaning procedures outlined in Section 3.2, the dataset was structured to allow an analysis of historical price changes and vendor-specific price patterns.
The dataset contains 1,996,969 rows and 5 columns, with variables capturing product (product name), vendor_name (name of the grocery retailer), price_current (current product price at the time of recording), price_old (previous recorded price for the product), and price_difference (difference between the current and old prices). Price data is recorded in Canadian dollars and captures a broad range of grocery items from various categories, including fresh produce, dairy, pantry staples, and household items.
To ensure the quality and consistency of the data, we focused on removing missing values and imputing outlier prices when necessary. For this analysis, we excluded entries with extreme price fluctuations beyond three standard deviations, as these could represent temporary discounts or data recording anomalies. This cleaning process provides a balanced and accurate reflection of grocery pricing, allowing for robust statistical analysis on vendor-specific trends and average price comparisons across product types.
By narrowing the scope to recent data, Project Hammer's dataset captures real-time fluctuations in grocery pricing, providing timely insights into potential market dynamics and consumer price impacts. The analysis allows for future investigation of temporal price trends, vendor-based price differences, and, more broadly, consumer affordability in a market facing rising cost-of-living pressures.

@@

Each entry in the data set represents a captured moment in time where prices, products, and vendor information were recorded. The measurement process involved regular data collection from the online pricing systems of the following prominent grocery chains: Voila, T&T, Loblaws, No Frills, Metro, Galleria, Walmart, and Save-On-Foods. Each entry captures variables such as:  product name, vendor name, and both current and previous prices. The Project Hammer initiative leveraged web scraping technology to ensure a consistent and standardized entry format for all retailers, documenting prices at regular intervals. This approach ensures that price points reflect real-time data rather than historical estimates or annual averages, providing a close approximation of consumer experiences. The data set records the price of each grocery item in Canadian dollars, including items from broad categories such as fresh produce, dairy, pantry goods, and household essentials, to present a holistic view of grocery costs.

@@

Essentially, the real world phenomenon we observed was the website of each vendor with a listing of their products for that week, we turned this phenomenon into an entry within out data set with the use of a screen-scrape of the website UI. This means a HTTP request to load the page was sent to the desired vendors website on a specific day, than the returned HTML was parsed to extract specific content such as text, images, prices, etc. These captured features were than utilized to fill out the corresponding columns within the data set.

@@

#### Read data ####
# Load the cleaned dataset
# analysis_data <- read_csv("C:/Users/eliza/Downloads/hammer-3-compressed/cleaned_price_changes.csv", show_col_types = FALSE)
# 
# # Generate plot for Average Current Price by Vendor
# vendor_price_plot <- analysis_data %>%
#   group_by(vendor_name) %>%
#   summarize(avg_price_current = mean(price_current, na.rm = TRUE)) %>%
#   ggplot(aes(x = reorder(vendor_name, avg_price_current), y = avg_price_current)) +
#   geom_bar(stat = "identity", fill = "skyblue") +
#   coord_flip() +
#   labs(
#     title = "Average Current Price by Vendor",
#     x = "Vendor",
#     y = "Average Current Price (CAD)"
#   ) +
#   theme_minimal()
# 
# # Display the plot
# vendor_price_plot

@@

# Average price difference by vendor
# price_diff_plot <- analysis_data %>%
#   group_by(vendor_name) %>%
#   summarize(avg_price_difference = mean(price_difference, na.rm = TRUE)) %>%
#   ggplot(aes(x = reorder(vendor_name, avg_price_difference), y = avg_price_difference)) +
#   geom_bar(stat = "identity", fill = "coral") +
#   coord_flip() +
#   labs(
#     title = "Average Price Difference by Vendor",
#     x = "Vendor",
#     y = "Average Price Difference (CAD)"
#   ) +
#   theme_minimal()
# price_diff_plot

@@

# Scatter plot of previous price vs. current price
# price_relationship_plot <- ggplot(analysis_data, aes(x = price_old, y = price_current)) +
#   geom_point(alpha = 0.4, color = "blue") +
#   geom_smooth(method = "lm", color = "red") +
#   labs(
#     title = "Relationship Between Previous and Current Prices",
#     x = "Previous Price (CAD)",
#     y = "Current Price (CAD)"
#   ) +
#   theme_minimal()
# price_relationship_plot

@@

# Calculate mean and standard deviation of price_difference
# mean_diff <- mean(analysis_data$price_difference, na.rm = TRUE)
# sd_diff <- sd(analysis_data$price_difference, na.rm = TRUE)
# 
# # Filter out extreme values beyond 3 standard deviations
# filtered_data <- analysis_data %>%
#   filter(price_difference >= (mean_diff - 3 * sd_diff),
#          price_difference <= (mean_diff + 3 * sd_diff))
# 
# # Plot histogram with filtered data
# price_difference_hist <- ggplot(filtered_data, aes(x = price_difference)) +
#   geom_histogram(bins = 30, fill = "lightgreen", color = "black") +
#   labs(
#     title = "Distribution of Price Differences (Filtered)",
#     x = "Price Difference (CAD)",
#     y = "Frequency"
#   ) +
#   theme_minimal()
# price_difference_hist

@@

Table 1: Model Evaluation Metrics. The table displays the MAE and RMSE for the model, reflecting the average and squared prediction errors, respectively. Lower values indicate better predictive accuracy, suggesting that the model effectively captures the relationship between previous prices, vendor identities, and price differences.

@@

# Discussion {sec-discussion}

@@

\appendix

# Appendix {-}

>>>>

---
title: "Analyzing changes in Canadian Grocery Prices"
# subtitle: "My subtitle if needed"
author: 
  - Abdullah Motasim
  - Elizabeth Luong
thanks: "Code and data are available at: [https://github.com/Luongel1/project_hammer](https://github.com/Luongel1/project_hammer)."
date: today
date-format: long
abstract: "This paper investigates grocery pricing trends across Canada using SQL analysis on the 'Project Hammer' dataset, which contains detailed vendor-specific price data. Focusing on statistical relationships and potential pricing biases, we explore factors driving price variations across different regions and vendors. A Bayesian regression model is applied to understand correlations, accounting for limitations and highlighting areas where causation cannot be assumed. Findings reveal competitive dynamics within Canadian grocery markets, implications for consumers, and recommendations for future research into market and pricing behavior across vendors."
format: pdf
number-sections: true
bibliography: references.bib
---

@@

library(tidyverse)
library(ggplot2)
library(dplyr)
# library(knitr)
# library(rstanarm)
# library(arrow)
# library(data.table)
library(here)

analysis_data <- read_csv(here("data/02-analysis_data/cleaned_price_changes.csv"), show_col_types = FALSE)
```

@@

The rising cost of groceries in Canada has drawn significant public attention, impacting consumer budgets and prompting regulatory interest. While inflation and supply disruptions are often cited as contributors, the specifics of how major Canadian grocery vendors set and adjust prices remain underexplored. This study addresses this gap by analyzing data from Project Hammer, a dataset that tracks grocery prices across vendors including Voila, T&T, Loblaws, No Frills, Metro, Galleria, Walmart, and Save-On-Foods. Through examining these data, we aim to uncover factors driving grocery price dynamics and assess the degree to which vendor-specific pricing may reflect competitive or non-competitive practices.

@@

The paper is structured as follows: @sec-data discusses the data types included in the raw data, the cleaning process for the data, and the reason for selecting the dataset we did. @sec-model discusses model specification and justification for utilizing a Bayesian linear regression model. @sec-results presents the trends and correlations between different variables utilizing tabular and graphical means. @sec-discussion discusses the results of @sec-results going into detail on what the simulation results can tell us about grocery prices in Canada as well as discussing missing data and sources of bias.

@@

For this study, we utilize data from Project Hammer, a Canadian initiative designed to monitor grocery prices across major retailers in Canada. The Project Hammer dataset was collected from eight prominent Canadian grocery vendors—Voila, T&T, Loblaws, No Frills, Metro, Galleria, Walmart, and Save-On-Foods—between February 28, 2024, and the latest available data load in the database [@hammer]. This dataset enables us to investigate price fluctuations, identify pricing trends, and explore potential competitive or collusive patterns within the Canadian grocery sector.
Project Hammer aims to support regulatory efforts by providing data on price transparency, making it suitable for academic research and potential legal analysis. Following data processing and cleaning procedures outlined in Section 3.2, the dataset was structured to allow an analysis of historical price changes and vendor-specific price patterns.
The dataset contains 1,996,969 rows and 5 columns, with variables capturing product (product name), vendor_name (name of the grocery retailer), price_current (current product price at the time of recording), price_old (previous recorded price for the product), and price_difference (difference between the current and old prices). Price data is recorded in Canadian dollars and captures a broad range of grocery items from various categories, including fresh produce, dairy, pantry staples, and household items.
To ensure the quality and consistency of the data, we focused on removing missing values and imputing outlier prices when necessary. For this analysis, we excluded entries with extreme price fluctuations beyond three standard deviations, as these could represent temporary discounts or data recording anomalies. This cleaning process provides a balanced and accurate reflection of grocery pricing, allowing for robust statistical analysis on vendor-specific trends and average price comparisons across product types.
By narrowing the scope to recent data, Project Hammer's dataset captures real-time fluctuations in grocery pricing, providing timely insights into potential market dynamics and consumer price impacts. The analysis allows for future investigation of temporal price trends, vendor-based price differences, and, more broadly, consumer affordability in a market facing rising cost-of-living pressures.

@@

Each entry in the dataset represents a captured moment in time where prices, products, and vendor information were recorded. The measurement process involved regular data collection from the online pricing systems of the following prominent grocery chains: Voila, T&T, Loblaws, No Frills, Metro, Galleria, Walmart, and Save-On-Foods. Each entry captures variables such as:  product name, vendor name, and both current and previous prices. The Project Hammer initiative leveraged web scraping technology to ensure a consistent and standardized entry format for all retailers, documenting prices at regular intervals. This approach ensures that price points reflect real-time data rather than historical estimates or annual averages, providing a close approximation of consumer experiences. The dataset records the price of each grocery item in Canadian dollars, including items from broad categories such as fresh produce, dairy, pantry goods, and household essentials, to present a holistic view of grocery costs.

@@

Essentially, the real world phenomenon we observed was the website of each vendor with a listing of their products for that week, we turned this phenomenon into an entry within out dataset with the use of a screen-scrape of the website UI. This means a HTTP request to load the page was sent to the desired vendors website on a specific day, than the returned HTML was parsed to extract specific content such as text, images, prices, etc. These captured features were than utilized to fill out the corresponding columns within the dataset.

@@

# Generate plot for Average Current Price by Vendor
vendor_price_plot <- analysis_data %>%
  group_by(vendor_name) %>%
  summarize(avg_price_current = mean(price_current, na.rm = TRUE)) %>%
  ggplot(aes(x = reorder(vendor_name, avg_price_current), y = avg_price_current)) +
  geom_bar(stat = "identity", fill = "skyblue") +
  coord_flip() +
  labs(
    title = "Average Current Price by Vendor",
    x = "Vendor",
    y = "Average Current Price (CAD)"
  ) +
  theme_minimal()

# Display the plot
vendor_price_plot

@@

# Average price difference by vendor
price_diff_plot <- analysis_data %>%
  group_by(vendor_name) %>%
  summarize(avg_price_difference = mean(price_difference, na.rm = TRUE)) %>%
  ggplot(aes(x = reorder(vendor_name, avg_price_difference), y = avg_price_difference)) +
  geom_bar(stat = "identity", fill = "coral") +
  coord_flip() +
  labs(
    title = "Average Price Difference by Vendor",
    x = "Vendor",
    y = "Average Price Difference (CAD)"
  ) +
  theme_minimal()
price_diff_plot

@@

# Scatter plot of previous price vs. current price
price_relationship_plot <- ggplot(analysis_data[0:10000,], aes(x = price_old, y = price_current)) +
  geom_point(alpha = 0.4, color = "blue") +
  geom_smooth(method = "lm", color = "red") +
  labs(
    title = "Relationship Between Previous and Current Prices",
    x = "Previous Price (CAD)",
    y = "Current Price (CAD)"
  ) +
  theme_minimal()
price_relationship_plot
# price_time_plot <- ggplot(analysis_data, aes(x = price_old)) +
#   # geom_line(aes(y = price_old, color = "Previous Price"), size = 1) +
#   geom_line(aes(y = price_current, color = "Current Price"), size = 1) +
#   labs(
#     title = "Price Changes Over Time",
#     x = "Old price",
#     y = "New Price (CAD)",
#     color = "Price Type"
#   ) +
#   theme_minimal()
# price_time_plot
# 
# max_price_old <- max(analysis_data$price_old, na.rm = TRUE)
# max_price_old

@@

# Calculate mean and standard deviation of price_difference
mean_diff <- mean(analysis_data$price_difference, na.rm = TRUE)
sd_diff <- sd(analysis_data$price_difference, na.rm = TRUE)

# Filter out extreme values beyond 3 standard deviations
filtered_data <- analysis_data %>%
  filter(price_difference >= (mean_diff - 3 * sd_diff),
         price_difference <= (mean_diff + 3 * sd_diff))

# Plot histogram with filtered data
price_difference_hist <- ggplot(filtered_data, aes(x = price_difference)) +
  geom_histogram(bins = 30, fill = "lightgreen", color = "black") +
  labs(
    title = "Distribution of Price Differences (Filtered)",
    x = "Price Difference (CAD)",
    y = "Frequency"
  ) +
  theme_minimal()
price_difference_hist

@@

<!-- Table 1: Model Evaluation Metrics. The table displays the MAE and RMSE for the model, reflecting the average and squared prediction errors, respectively. Lower values indicate better predictive accuracy, suggesting that the model effectively captures the relationship between previous prices, vendor identities, and price differences. -->

@@

# Discussion {#sec-discussion}

@@
